---
title: "L.A. homeless arrests analysis"
author: "Christine Zhang"
output: pdf_document
---

```{r, message=FALSE, warning=FALSE}
#install.packages('dplyr','feather','ggplot2')
library('dplyr')
library('feather')
library('ggplot2')
```

```{r}
unzip("arrests.zip")
```

Load the data:
This is a feather filetype but you can load whatever file you like.
For example, I often read CSV files into R. I would write something like this: read_csv("~/Documents/folder/blah.csv")
```{r}
data <- read_feather('arrests.feather')
```

```{r}
# view column headers
names(data)
```

```{r}
# view first few rows of data set
head(data)
```
# Finding 1: The LAPD made 14,000 arrests of homeless people last year, a 30% increase over 2011

Group the data by arrest year and homeless and sum the total number of arrests
```{r}
arrest.totals <- data %>% 
  group_by(arrest_year, homeless) %>% 
  distinct(booking_num) %>% 
  summarize(arrests_number = n())
```

Filter to homeless arrests
```{r}
homeless.totals <- arrest.totals %>% filter(homeless == 1)
homeless.totals
```

```{r}
print(paste0("The *raw* increase in homeless arrests between 2011 and 2016 is ", 
            round((homeless.totals[homeless.totals$arrest_year == 2016,]$arrests_number / 
            homeless.totals[homeless.totals$arrest_year == 2011,]$arrests_number - 1) * 100), "%"))
```

Fix the missing data issue:
```{r}
# extract the booking dates as a vector
booking.dates <- data %>% select(booking_ymd)
booking.dates <- booking.dates %>% distinct(booking_ymd) %>% select(booking_ymd) %>% arrange(booking_ymd)
booking.dates$has.data = 1

# get the time period (minimum date and maximum date) of the data set
time.min <- booking.dates$booking_ymd[1]
time.max <- booking.dates$booking_ymd[length(booking.dates$booking_ymd) - 1] 

# create a dataframe of all the days spanning that time period
all.dates.frame <- data.frame(list(booking_ymd = seq(time.min, time.max, by="day")))

# merge this dataframe with the vector of booking dates to find the missing dates
merged.data <- merge(all.dates.frame, booking.dates , all=T)
missing.dates <- merged.data %>% filter(is.na(has.data) == T)
```

```{r}
#view missing.dates
missing.dates
```

Pro-rate the 2011 figure to account for the missing six days
```{r}
prorated.homeless.2011 <- homeless.totals[homeless.totals$arrest_year == 2011,]$arrests_number/(365 - 6) * 365
prorated.homeless.2011
```

```{r}
print(paste0("The *prorated* change in homeless arrests between 2011 and 2016 is ", 
            round((homeless.totals[homeless.totals$arrest_year == 2016,]$arrests_number / 
            prorated.homeless.2011 - 1) * 100), "% (rounded down to 30% in the story)"))
```


# Finding 2: LAPD arrests overall went down 15% from 2011 to 2016
```{r}
all.totals <- arrest.totals %>% 
  group_by(arrest_year) %>% 
  summarize(arrests_number = sum(arrests_number))
```

```{r}
print(all.totals)
```

```{r}
print(paste0("The *raw* change in overall arrests between 2011 and 2016 is ", 
            round((all.totals[all.totals$arrest_year == 2016,]$arrests_number / 
            all.totals[all.totals$arrest_year == 2011,]$arrests_number - 1) * 100), "%"))
```

```{r}
# Again, we need to pro-rate to take into account the six missing days in 2011.
prorated.arrests.2011 <- all.totals[all.totals$arrest_year == 2011,]$arrests_number/(365 - 6) * 365
prorated.arrests.2011

```

```{r}
print(paste0("The *prorated* change between 2011 and 2016 is ", 
            round((all.totals[all.totals$arrest_year == 2016,]$arrests_number / 
            prorated.arrests.2011 - 1) * 100), "%"))
```


#Finding 3: Two-thirds of those arrested were black or Latino

```{r}
arrests.race <- data %>% 
  group_by(arrest_year, homeless, race)  %>%
  distinct(booking_num)
```

```{r}
# Create a variable, race.grp to represent racial/ethnic grouping, where W = white, B = black, H = Latino (Hispanic), etc.

table(arrests.race$race)
```


```{r}
#rename variables
arrests.race$race.grp <- ifelse(arrests.race$race == 'W', "White",
                                ifelse(arrests.race$race == 'B', "Black",
                                       ifelse(arrests.race$race == 'H', "Latino",
                                              ifelse(arrests.race$race == 'A' | arrests.race$race == 'C' | arrests.race$race == 'J'| arrests.race$race == 'K'| arrests.race$race == 'F', "Asian",
                                                     'Other'))))
```
```{r}
#Group by race.grp and calculate the total number and percentage of homeless arrests
arrests.race.yr <- arrests.race %>%
  group_by(arrest_year, homeless, race.grp) %>%
  summarize(arrests_number = n()) %>%
  mutate(arrests_percent = arrests_number / sum(arrests_number) * 100) 

arrests.race.yr %>% filter(homeless == 1 & arrest_year == 2016) %>% arrange(desc(arrests_percent))
```
```{r}
ggplot(arrests.race.yr %>% filter(homeless == 1 & arrest_year != 2017), aes(x = arrest_year, 
                                                                            y = arrests_percent, color = race.grp)) +
  geom_line() +
  geom_text(data = arrests.race.yr %>% filter(homeless == 1 & 
                                              arrest_year == 2016), aes(label = race.grp), hjust = 0.7,
            vjust = 1) +
  scale_y_continuous(limits = c(0, 50)) +
  labs(x = "", y = "% of arrests", title = "Racial Breakdown of homeless arrests") +
  theme(legend.position = 'none')
```
# Finding 4: In 2011, one in 10 people arrests citywide were of homeless people; in 2016, it was 1 in 6
```{r}
# Use the grouped dataframe arrest.totals to calculate percentage of homeless arrests by year
arrest.totals %>% 
    mutate(arrests_percent = arrests_number / sum(arrests_number) * 100) %>%
    filter(homeless == 1)
```

# Finding 5: The 14,000 arrests of homeless people in 2016 included more than 500 unique charges

```{r}
#Filter the data to include homeless arrests in 2016 and calculate the number and percent of times each charge was cited. Note that this is done by analyzing each charge separately, so the times_cited column will not sum to the total number of arrests per year (arrestees can have multiple charges).

arrest.reasons <- data %>% filter(homeless == 1 & arrest_year == 2016) %>% 
                           group_by(charge_code, charge_desc) %>% 
                           summarize(times_cited = n()) %>% 
                           ungroup() %>% 
                           mutate(percent_cited = times_cited/sum(times_cited) * 100)
```

```{r}
#get the number of unique charges 
length(unique(arrest.reasons$charge_code))
```
#Finding 6: The most common offense was failure to appear in court for unpaid petty or minor citations
```{r}
#Sort by percent of the time the charge was cited to get the top charges

head(arrest.reasons %>% arrange(desc(percent_cited)))
```
Many codes did not come with charge descriptions in the data. Those that appear in the above table are described as follows:

* 459.5PC: shoplifting
* 3000.08CPC: parole warrant
* 3454(C)PC: flash incarceration

#Finding 7: The top five charges were for non-violent or minor offenses

Some charge codes are grouped. For example. charge codes 40508(A)VC, 853.7PC, and 853.8PC all cover failure to appear.

```{r}
arrest.reasons$failure <- ifelse(arrest.reasons$charge_code == '40508(A)VC'|
                                 arrest.reasons$charge_code == '853.7PC'|
                                 arrest.reasons$charge_code == '853.8PC', 1, 0)

arrest.reasons$trespass <- ifelse(arrest.reasons$charge_code == '419PC'|
                                  arrest.reasons$charge_code == '602(K)PC'|
                                  arrest.reasons$charge_code == '602(O)(2)PC'|
                                  arrest.reasons$charge_code == '602.5(A)PC'|
                                  arrest.reasons$charge_code == '555PC'|
                                  arrest.reasons$charge_code == '484F(A)PC'|
                                  arrest.reasons$charge_code == '602(L)(1)PC'|
                                  arrest.reasons$charge_code == '602(P)PC'|
                                  arrest.reasons$charge_code == '602.5(B)PC'|
                                  arrest.reasons$charge_code == '602PC'|
                                  arrest.reasons$charge_code == '602(M)PC'|
                                  arrest.reasons$charge_code == '602(Q)PC'|
                                  arrest.reasons$charge_code == '602.8PC'|
                                  arrest.reasons$charge_code == '602(A)PC'|
                                  arrest.reasons$charge_code == 'A602(N)1PC'|
                                  arrest.reasons$charge_code == '602(S)PC'|
                                  arrest.reasons$charge_code == '626.8(A)1PC'|
                                  arrest.reasons$charge_code == '602(D)PC'|
                                  arrest.reasons$charge_code == '602(N)PC'|
                                  arrest.reasons$charge_code == '602(U)(1)PC'|
                                  arrest.reasons$charge_code == '647(E)PC'|
                                  arrest.reasons$charge_code == '602(F)PC'|
                                  arrest.reasons$charge_code == '602(O)PC'|
                                  arrest.reasons$charge_code == '602.1(A)PC'| 
                                  arrest.reasons$charge_code == '647(H)PCLPP'| 
                                  arrest.reasons$charge_code == '602(J)PC'| 
                                  arrest.reasons$charge_code == '602(O)(1)PC'| 
                                  arrest.reasons$charge_code == '602.1(B)PC'| 
                                  arrest.reasons$charge_code == '369I(A)PC', 1, 0)

arrest.reasons$shoplift <- ifelse(arrest.reasons$charge_code == '18 1708'|
                                  arrest.reasons$charge_code == '484PCTFMV'|
                                  arrest.reasons$charge_code == '485PC'|
                                  arrest.reasons$charge_code == '488PC'|
                                  arrest.reasons$charge_code == '459.5PC'|
                                  arrest.reasons$charge_code == '484F(A)PC'|
                                  arrest.reasons$charge_code == 'A488PC'|
                                  arrest.reasons$charge_code == '490PC'|
                                  arrest.reasons$charge_code == 'A484PC'|
                                  arrest.reasons$charge_code == '484E(D)PC'|
                                  arrest.reasons$charge_code == '666PC'|
                                  arrest.reasons$charge_code == '484PC'|
                                  arrest.reasons$charge_code == '490.2PC'|
                                  arrest.reasons$charge_code == '666(A)PC'|
                                  arrest.reasons$charge_code == '484(A)PC'|
                                  arrest.reasons$charge_code == '490.5(A)PC'|
                                  arrest.reasons$charge_code == '537(A)(1)PC'|
                                  arrest.reasons$charge_code == '666.5PC'|
                                  arrest.reasons$charge_code == '484E(A)PC'|
                                  arrest.reasons$charge_code == '587CPC'|
                                  arrest.reasons$charge_code == '666.5(A)PC'|
                                  arrest.reasons$charge_code == '484E(B)PC', 1, 0)

arrest.reasons$supervision_viol <- ifelse(arrest.reasons$charge_code == '1203.2PC'|
                                          arrest.reasons$charge_code == '3000.08CPC'|
                                          arrest.reasons$charge_code == '3454(C)PC'| 
                                          arrest.reasons$charge_code == '3455(B)1PC'|
                                          arrest.reasons$charge_code == '1203.2(A)PC'|
                                          arrest.reasons$charge_code == '3056PC'|
                                          arrest.reasons$charge_code == '3455(A)4PC'|
                                          arrest.reasons$charge_code == '3455(C)PC'|
                                          arrest.reasons$charge_code == '3000.08FPC'|
                                          arrest.reasons$charge_code == '3454PC'|
                                          arrest.reasons$charge_code == '3455(A)PC'|
                                          arrest.reasons$charge_code == '18 3606US', 1, 0)

arrest.reasons$drug_poss <- ifelse(arrest.reasons$charge_code == '11377(A)HS'|
                                   arrest.reasons$charge_code == '11377(A)1HS'|
                                   arrest.reasons$charge_code == '11377HS'| 
                                   arrest.reasons$charge_code == '11350(A)HS'|
                                   arrest.reasons$charge_code == '11350HS'|
                                   arrest.reasons$charge_code == '11357HS'|
                                   arrest.reasons$charge_code == '11357(A)HS'|
                                   arrest.reasons$charge_code == '11357(B)HS'|
                                   arrest.reasons$charge_code == '11357(C)HS'|
                                   arrest.reasons$charge_code == '4573.6PC'|
                                   arrest.reasons$charge_code == '11550(A)HS'|
                                   arrest.reasons$charge_code == '11375(B)2HS'|
                                   arrest.reasons$charge_code == '11351HS'|
                                   arrest.reasons$charge_code == '4060BP', 1, 0)

arrest.reasons$charge_desc_grouped <- ifelse(arrest.reasons$drug_poss == 1, 'drug_poss',
                                             ifelse(arrest.reasons$trespass == 1, 'trespass',
                                                    ifelse(arrest.reasons$shoplift == 1, 'shoplift',
                                                           ifelse(arrest.reasons$supervision_viol == 1, 'supervision violation',
                                                                  ifelse(arrest.reasons$failure == 1, 'failure to appear',
                                                                         arrest.reasons$charge_code)))))
```


Get top five offenses using charge_desc_grouped as the charge identifier
```{r}
arrest.reasons %>% group_by(charge_desc_grouped) %>% 
    summarise(times_cited = sum(times_cited)) %>% 
    mutate(perc_cited = times_cited/sum(times_cited) * 100) %>% arrange(desc(times_cited)) %>% head(5)
```

